---
title: "Analyzing Toronto Apartment Rental Prices"
author:
- "Austin Hart (45726379)"
- "Olivia Jiang (82785783)"
- "Woobin Kim (39017496)"
- "Raina Tian (11946720)"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 7)
```

Project Task Assignments:

- Austin did...
- Olivia did...
- Raina did...
- Woobin did...

\newpage

## Introduction

This section should describe the background and objectives of the study.

## Data Description

<!-- This section should describe where the dataset comes from, the sample size, how the data were collected, etc. -->
We obtained the data from [\underline{this}](<https://www.kaggle.com/datasets/rajacsp/toronto-apartment-price>) Kaggle dataset. The amount of rows in the original dataset before any data cleaning is 1,124. The data contains information on apartment rental units in Toronto collected in 2018, including the apartment's geographic information (`address`, `latitude`, and `longitude`), information on its rooms (the number of `bedroom`s, `bathroom`s, and `den`s), and the unit's monthly rental `price`.

The total number of rows in the original dataset is 1,124. There are no missing values in any columns.

During data cleaning, we removed 308 duplicates which likely occurred because the data came from multiple sources. We also removed 5 rows with unusual values for `price` (\$65/month, \$99/month, \$99/month, \$36,900/month, \$535,000/month). These outliers may skew our results so we decided to remove them. The sample size is now `n = 811`.

## Exploratory Data Analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
# Loading packages
library(tidyverse)
library(knitr)
library(scales)
library(patchwork)
library(corrplot)
library(broom)

# Reading the data
data <- read_csv("Toronto_apartment_rentals_2018.csv")

nrow(data)

# Seeing how many duplicate rows there are
sum(duplicated(data)) # 308

# Viewing duplicate rows
data[duplicated(data) | duplicated(data, fromLast = TRUE), ]

# Cleaning the data
data <- data %>%
  # Removing duplicate rows
  distinct() %>%
  # Removing unnecessary columns
  select(-c(Address)) %>%
  # Making the column names lowercase
  rename_with(tolower) %>%
  # Converting price to a number
  mutate(price = parse_number(price)) %>%
  # Removing outliers
  filter(!price %in% c(65, 99, 99, 36900, 535000))

# Seeing if there are any NAs
colSums(is.na(data)) # 0

# Adding a binary response variable (whether price is more than its mean)
data <- data %>%
  mutate(high_price = if_else(price > mean(price), 1, 0))
```

<!-- This section may include summary statistics (e.g., means and standard deviations of the variables), graphical displays of the data, and any preliminary conclusions. No models are assumed in this section. -->

Before modeling, we want to find out more about the data. The variables we're most interested in are `price`, `bedroom`, `bathroom`, and `den`; however, we first view the sample's summary statistics of each column.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Viewing summary stats
data %>%
  summarise(across(where(is.numeric), list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)))) %>%
  tidy()
```

TODO: a table of summary stats of everything except address

To visualize the distribution of `price`, we plot a histogram.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Viewing a histogram of price
data %>%
  ggplot(aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "skyblue", bins = 20) +
  labs(title = "Distribution of Toronto Monthly Apartment Rents (2018)",
       x = "Monthly Rent",
       y = "Frequency") +
  geom_vline(aes(xintercept = mean(price)), 
             color = "black", linetype = "dashed", linewidth = 0.5) +
  annotate("text", 
           x = 2255 + 200,
           y = 200,
           label = "Mean Rent: $2,186", 
           color = "black", 
           hjust = 0) +
  scale_x_continuous(labels = label_dollar()) +
  theme_classic()
```

The distribution is slightly right-skewed and unimodal around the mean of `mean`. As well, its standard deviation is \$1,021/month, the minimum is \$150/month, and the maximum is \$9,750/month.
 
TODO: more commentary
 
\newpage

Next, we view the frequencies of each discrete value of `bedroom`, `bathroom`, and `den`.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Making bar plots of bedroom/bathroom/den counts

bar_1 <- data %>%
  ggplot(aes(x = bedroom)) +
  geom_bar(fill = "darkolivegreen3") +
  labs(x = "# of Bedrooms", y = "Count") +
  coord_cartesian(ylim = c(0, 700)) +
  theme_classic()

bar_2 <- data %>%
  ggplot(aes(x = bathroom)) +
  geom_bar(fill = "darkolivegreen3") +
  labs(x = "# of Bathrooms", y = "Count") +
  coord_cartesian(ylim = c(0, 700)) +
  theme_classic()

bar_3 <- data %>%
  mutate(den = as.factor(den)) %>%
  ggplot(aes(x = den)) +
  geom_bar(fill = "darkolivegreen3") +
  labs(x = "# of Dens", y = "Count") +
  coord_cartesian(ylim = c(0, 700)) +
  theme_classic()

bar_1 + bar_2 + bar_3 + plot_annotation(title = "Frequencies of Room Counts by Room Type")
```

TODO: comment on them (such as the unbalanced groups)

Below are boxplots of `price` split by `bedroom`, `bathroom`, and `den` counts in order to see how `price`'s distribution varies by each of these 3 apartment characteristics.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Making boxplots of price by bed/bath/den count

boxplot_1 <- data %>%
  mutate(bedroom = as.factor(bedroom)) %>%
  ggplot(aes(x = bedroom, y = price)) +
  geom_boxplot(fill = "lightgrey") +
  labs(y = "Monthly Rent", x = "# of Bedrooms") +
  scale_y_continuous(labels = label_dollar()) +
  theme_classic()

boxplot_2 <- data %>%
  mutate(bathroom = as.factor(bathroom)) %>%
  ggplot(aes(x = bathroom, y = price)) +
  geom_boxplot(fill = "lightgrey") +
  labs(y = "Monthly Rent", x = "# of Bathrooms") +
  scale_y_continuous(labels = label_dollar()) +
  theme_classic()

boxplot_3 <- data %>%
  mutate(den = as.factor(den)) %>%
  ggplot(aes(x = den, y = price)) +
  geom_boxplot(fill = "lightgrey") +
  labs(y = "Monthly Rent", x = "# of Dens") +
  scale_y_continuous(labels = label_dollar()) +
  theme_classic()

boxplot_1 + boxplot_2 + boxplot_3 + plot_annotation(title = "Distributions of Rent Prices by Room Counts")
```

TODO: comment on them

TODO: number of observations of each 0/1 column

\newpage

We're also curious about the correlations between the variables so we plot (the upper triangle of) a correlation matrix.

```{r, echo=FALSE, warnings=FALSE, fig.height=3, fig.width=5, fig.align='center'}
# Making a correlation matrix
corrplot(round(cor(select(data, -high_price)), 2),
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 30,
         tl.cex = 0.8,
         cl.ratio = 0.35,
         cl.length = 5,
         cl.cex = 0.9,
         number.cex = 0.8,
         col = colorRampPalette(c("#4477AA", "#77AADD", "#FFFFFF", "#EE9988", "#BB4444"))(200),
         mar = c(2,2,2,2),
         diag = TRUE)
```
The variables with the strongest correlation with `price` are `bathroom` (r = 0.47) and `bathroom` (r = 0.38), though these are only moderately strong linear correlations. `den` is weakly negatively correlated with `price` at r = -0.01. Moreoever, `bedroom` and `bathroom` are moderately correlated at r = 0.50, which may indicate that they're collinear, which may affect our modeling decisions later on.

TODO: key take-aways of EDA

\newpage

## Models

<!-- This section should include at least two statistical models for the same dataset, justifications of the choices of the models, model diagnostics, parameter estimates and their standard errors and p-values, and interpretation of the results. -->

### Additive Linear Regression

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Linear additive model

linear_add <- lm(price ~ bedroom + bathroom, data = data)

# summary(linear_add)

# Coefficients & p-values
# TODO: comments on p-values/significance, interpretations of coefficients in the context of our data
tidy(linear_add, conf.int = TRUE, conf.level = 0.95) %>%
  mutate_if(is.numeric, round, 2) %>%
  kable()
```

TODO: commentary

```{r, echo=FALSE, fig.height=6}
# Residual plot
plot(linear_add, which = 1)
```

TODO: commentary

```{r, echo=FALSE, fig.height=6}
# Q-Q plot
plot(linear_add, which = 2)
```

### Additive Logistic Regression

```{r, echo=FALSE}
# Additive logistic model

logistic_add <- glm(high_price ~ bedroom + bathroom, data = data)

# summary(logistic_add)

# Coefficients & p-values
tidy(logistic_add, conf.int = TRUE, conf.level = 0.95) %>%
  mutate_if(is.numeric, round, 2) %>%
  kable()
```

TODO: commentary

TODO: maybe precision, accuracy, etc. as diagnostics

## Conclusion

This section should provide overall conclusions, any limitations in the datasets and analyses, and validity of the conclusions.

\newpage

## Appendix

```{r, warning=FALSE, message=FALSE}
# Loading packages
library(tidyverse)
library(knitr)
library(scales)
library(patchwork)
library(corrplot)
library(broom)
```

