---
title: "Modeling Toronto Apartment Rental Prices"
author: "Austin Hart, Olivia Jiang, Woobin Kim, Raina Tian"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This section should describe the background and objectives of the study.

## Data Description

<!-- This section should describe where the dataset comes from, the sample size, how the data were collected, etc. -->
We obtained the data from <https://www.kaggle.com/datasets/rajacsp/toronto-apartment-price>. The total number of rows in the original dataset before any data cleaning is 1,124. The data is made up of rental prices and contains data on apartment rental units in Toronto collected in 2018. It contains the apartment's geographic information (address, latitude, longitude), information on the rooms in each unit (the number of bedrooms, bathrooms, and dens), and the unit's monthly rental cost in CAD.

## Exploratory Data Analysis (EDA)

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
# Loading packages
library(tidyverse)

# Reading the data
data <- read_csv("Toronto_apartment_rentals_2018.csv")

# Seeing how many duplicate rows there are
sum(duplicated(data)) # 308

# Viewing duplicate rows
data[duplicated(data) | duplicated(data, fromLast = TRUE), ]

# Cleaning the data
data <- data %>%
  # Removing unnecessary columns
  select(-c(Address, Lat, Long)) %>%
  # Making the column names lowercase
  rename_with(tolower) %>%
  # Converting price to a number
  mutate(price = parse_number(price)) %>%
  # Removing outliers
  # NOTE: there's also two 99s I don't know if we should keep
  filter(!price %in% c(65, 99, 99, 36900, 535000)) %>%
  # Removing duplicate rows
  distinct()

# Seeing if there are any NAs
colSums(is.na(data)) # 0
```

<!-- This section may include summary statistics (e.g., means and standard deviations of the variables), graphical displays of the data, and any preliminary conclusions. No models are assumed in this section. -->

The total number of rows in the original dataset is 1,124. During data cleaning, we removed 308 duplicates. We also removed 5 rows with unusual values for `price` (\$65, \$99, \$99, \$36,900, \$535,000). These outliers may skew our results so we decided to remove them. The sample size is now `n = 811`. There are no missing values in any columns.

Before modeling, we want to find out more about the data. The variables we're most interested in are `price`, `bedroom`, `bathroom`, and `den`.

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
# Viewing summary stats of price
mean(data$price) # 2254.844
sd(data$price) # 1020.881
min(data$price) # 150
max(data$price) # 9750
```

This histogram of `price` shows that the distribution is slightly right-skewed and unimodal around the mean of \$2255. As well, its standard deviation is \$1021, the minimum is \$150, and the maximum is \$9750.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Viewing a histogram of price
data %>%
  ggplot(aes(x = price)) +
  geom_histogram(fill = "skyblue", bins = 20) +
  labs(x = "Monthly Rental Cost (2018)", y = "Frequency") +
  geom_vline(aes(xintercept = mean(price)), 
             color = "black", linetype = "dashed", linewidth = 1) +
  annotate("text", 
           x = 2255 + 200,
           y = 80,
           label = "Mean Monthly Rent: $2255", 
           color = "black", 
           hjust = 0) +
  theme_minimal()
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Viewing frequencies of bed/bath/den
# TODO: put these in proper tables
table(data$bedroom) # 3 levels
table(data$bathroom) # 5 levels
table(data$den) # 2 levels
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Making boxplots of price by bed/bath/den count
# TODO: label axes, put title, make pretty

data %>%
  mutate(bedroom = as.factor(bedroom)) %>%
  ggplot(aes(x = bedroom, y = price)) +
  geom_boxplot()

data %>%
  mutate(bathroom = as.factor(bathroom)) %>%
  ggplot(aes(x = bathroom, y = price)) +
  geom_boxplot()

data %>%
  mutate(den = as.factor(den)) %>%
  ggplot(aes(x = den, y = price)) +
  geom_boxplot()
```

```{r, echo=FALSE}
# Making scatterplots of price against bedroom and bathroom 
# NOTE: these relationships don't seem very linear

data %>%
  ggplot(aes(x = bedroom, y = price)) +
  geom_point()

data %>%
  ggplot(aes(x = bedroom, y = price)) +
  geom_point()
```

```{r, echo=FALSE}
# Making a correlation matrix
# NOTE: bed and bath are moderately correlated at r = 0.6365081
# NOTE: price and den aren't really correlated at r = 0.03690185
# TODO: make pretty by making it into a proper figure
cor(data)
```

```{r, echo=FALSE}
# Making histograms
# hist(data$Bedroom) # TODO - only integers on x-axis
# hist(data$Bathroom)
# hist(data$Den)
# hist(data$Price)
# data[order(data$Price, decreasing = FALSE),]
```

## Models

This section should include at least two statistical models for the same dataset, justifications of the choices of the models, model diagnostics, parameter estimates and their standard errors and p-values, and interpretation of the results.

```{r, echo=FALSE}
model_1 <- lm(price ~ ., data)
summary(model_1)
```

## Conclusion

This section should provide overall conclusions, any limitations in the datasets and analyses, and validity of the conclusions.

## References

Any references can go here.

## Appendix

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

