---
title: "Modeling Toronto Apartment Rental Prices"
author: "Austin Hart, Olivia Jiang, Woobin Kim, Raina Tian"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 7)
```

\newpage

## Introduction

This section should describe the background and objectives of the study.

## Data Description

<!-- This section should describe where the dataset comes from, the sample size, how the data were collected, etc. -->
We obtained the data from [\underline{this}](<https://www.kaggle.com/datasets/rajacsp/toronto-apartment-price>) Kaggle dataset. The amount of rows in the original dataset before any data cleaning is 1,124. The data is made up of rental prices and contains data on apartment rental units in Toronto collected in 2018. It contains the apartment's geographic information (address, latitude, longitude), information on the rooms in each unit (the number of bedrooms, bathrooms, and dens), and the unit's monthly rental cost.

## Exploratory Data Analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
# Loading packages
library(tidyverse)
library(knitr)
library(scales)
library(patchwork)
library(corrplot)

# Reading the data
data <- read_csv("Toronto_apartment_rentals_2018.csv")

# Seeing how many duplicate rows there are
sum(duplicated(data)) # 308

# Viewing duplicate rows
data[duplicated(data) | duplicated(data, fromLast = TRUE), ]

# Cleaning the data
data <- data %>%
  # Removing unnecessary columns
  select(-c(Address, Lat, Long)) %>%
  # Making the column names lowercase
  rename_with(tolower) %>%
  # Converting price to a number
  mutate(price = parse_number(price)) %>%
  # Removing outliers
  filter(!price %in% c(65, 99, 99, 36900, 535000)) %>%
  # Removing duplicate rows
  distinct()

# Seeing if there are any NAs
colSums(is.na(data)) # 0
```

<!-- This section may include summary statistics (e.g., means and standard deviations of the variables), graphical displays of the data, and any preliminary conclusions. No models are assumed in this section. -->

The total number of rows in the original dataset is 1,124. During data cleaning, we removed 308 duplicates which likely occurred because the data came from multiple sources. We also removed 5 rows with unusual values for `price` (\$65/month, \$99/month, \$99/month, \$36,900/month, \$535,000/month). These outliers may skew our results so we decided to remove them. The sample size is now `n = 811`. There are no missing values in any columns.

Before modeling, we want to find out more about the data. The variables we're most interested in are `price`, `bedroom`, `bathroom`, and `den`.

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Viewing summary stats of price
mean(data$price) # 2254.844
sd(data$price) # 1020.881
min(data$price) # 150
max(data$price) # 9750
```

This histogram of `price` shows that the distribution is slightly right-skewed and unimodal around the mean of \$2255. As well, its standard deviation is \$1021, the minimum is \$150, and the maximum is \$9750. TODO: commentary

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Viewing a histogram of price
data %>%
  ggplot(aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "skyblue", bins = 20) +
  labs(title = "Distribution of Toronto Monthly Apartment Rents (2018)",
       x = "Monthly Rent",
       y = "Frequency") +
  geom_vline(aes(xintercept = mean(price)), 
             color = "black", linetype = "dashed", linewidth = 0.5) +
  annotate("text", 
           x = 2255 + 200,
           y = 80,
           label = "Mean Rent: $2,255", 
           color = "black", 
           hjust = 0) +
  scale_x_continuous(labels = label_dollar()) +
  theme_classic()
```

\newpage

Next, we view the frequencies of each discrete value of `bedroom`, `bathroom`, and `den`.

TODO: comment on them!

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=2}
# Making bar plots of bedroom/bathroom/den counts

bar_1 <- data %>%
  ggplot(aes(x = bedroom)) +
  geom_bar(fill = "darkolivegreen3") +
  labs(x = "# of Bedrooms", y = "Count") +
  coord_cartesian(ylim = c(0, 300)) +
  theme_classic()

bar_2 <- data %>%
  ggplot(aes(x = bathroom)) +
  geom_bar(fill = "darkolivegreen3") +
  labs(x = "# of Bathrooms", y = "Count") +
  coord_cartesian(ylim = c(0, 300)) +
  theme_classic()

bar_3 <- data %>%
  mutate(den = as.factor(den)) %>%
  ggplot(aes(x = den)) +
  geom_bar(fill = "darkolivegreen3") +
  labs(x = "# of Dens", y = "Count") +
  coord_cartesian(ylim = c(0, 300)) +
  theme_classic()

bar_1 + bar_2 + bar_3 + plot_annotation(title = "Frequencies of Room Counts by Room Type")
```

Below are boxplots of `price` split by `bedroom`, `bathroom`, and `den` counts in order to see how `price`'s distribution varies by each of these 3 apartment characteristics.

TODO: comment on them (such as the unbalanced groups)!

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Making boxplots of price by bed/bath/den count

boxplot_1 <- data %>%
  mutate(bedroom = as.factor(bedroom)) %>%
  ggplot(aes(x = bedroom, y = price)) +
  geom_boxplot(fill = "lightgrey") +
  labs(y = "Monthly Rent", x = "# of Bedrooms") +
  scale_y_continuous(labels = label_dollar()) +
  theme_classic()

boxplot_2 <- data %>%
  mutate(bathroom = as.factor(bathroom)) %>%
  ggplot(aes(x = bathroom, y = price)) +
  geom_boxplot(fill = "lightgrey") +
  labs(y = "Monthly Rent", x = "# of Bathrooms") +
  scale_y_continuous(labels = label_dollar()) +
  theme_classic()

boxplot_3 <- data %>%
  mutate(den = as.factor(den)) %>%
  ggplot(aes(x = den, y = price)) +
  geom_boxplot(fill = "lightgrey") +
  labs(y = "Monthly Rent", x = "# of Dens") +
  scale_y_continuous(labels = label_dollar()) +
  theme_classic()

boxplot_1 + boxplot_2 + boxplot_3 + plot_annotation(title = "Distributions of Rent Prices by Room Counts")
```

\newpage

We're also curious about the correlations between the variables so we plot (the upper triangle of) a correlation matrix.

```{r, echo=FALSE, warnings=FALSE, fig.height=3, fig.width=5, fig.align='center'}
# Making a correlation matrix
corrplot(round(cor(data), 2),
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 30,
         tl.cex = 0.8,
         cl.ratio = 0.35,
         cl.length = 5,
         cl.cex = 0.9,
         number.cex = 0.8,
         col = colorRampPalette(c("#4477AA", "#77AADD", "#FFFFFF", "#EE9988", "#BB4444"))(200),
         mar = c(2,2,2,2),
         diag = TRUE)
```
The variables with the strongest correlation with `price` are `bathroom` (r = 0.47) and `bathroom` (r = 0.38), though these are only moderately strong linear correlations. `den` is weakly negatively correlated with `price` at r = -0.01. Moreoever, `bedroom` and `bathroom` are moderately correlated at r = 0.50, which may indicate that they're collinear, which may affect our modeling decisions later on.

TODO: key take-aways of EDA

\newpage

## Models

This section should include at least two statistical models for the same dataset, justifications of the choices of the models, model diagnostics, parameter estimates and their standard errors and p-values, and interpretation of the results.

```{r, echo=FALSE}
model_1 <- lm(price ~ ., data)
summary(model_1)
```

```{r, echo=FALSE}
model_2 <- lm(price ~ bedroom + bathroom, data)
summary(model_2)
# TODO: model justifications
# TODO: residual plots, Q-Q plots, comments on p-values/significance, interpretations of coefficients in the context of our data
```

## Conclusion

This section should provide overall conclusions, any limitations in the datasets and analyses, and validity of the conclusions.

## Appendix

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

