---
title: "Modeling Toronto Apartment Rental Prices"
author: "Austin Hart, Olivia Jiang, Woobin Kim, Raina Tian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This section should describe the background and objectives of the study.

## Data Description

This section should describe where the dataset comes from, the sample size, how the data were collected, etc.

Our data was found on <https://www.kaggle.com/datasets/rajacsp/toronto-apartment-price>.

## Exploratory Data Analysis (EDA)

This section may include summary statistics (e.g., means and standard deviations of the variables), graphical displays of the data, and any preliminary conclusions. No models are assumed in this section.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Loading packages
library(tidyverse)

# Reading in the data
setwd("/Users/oliviajiang/Downloads/STAT 301/STAT 301 Project")
data <- read_csv("Toronto_apartment_rentals_2018.csv")
head(data)
```

## Models

This section should include at least two statistical models for the same dataset, justifications of the choices of the models, model diagnostics, parameter estimates and their standard errors and p-values, and interpretation of the results.

```{r, echo=FALSE}
# Code will be here but hidden
```

## Conclusion

This section should provide overall conclusions, any limitations in the datasets and analyses, and validity of the conclusions.

## References

Any references can go here.

## Appendix

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

